---
title: "NCAA Scoring Prediction Project"
author: "Jay Jang, Jason Qiu"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(broom)
library(dplyr)
library(car)
library(patchwork)
library(MASS)
library(corrplot)
```

# 1. LOAD DATA

```{r}
data <- read_csv("data/CollegeBasketballPlayers2022.csv")
```

# 2. CLEANING FOR ORIGINAL MODEL

```{r}
data_clean <- data %>%
  dplyr::select(pts, yr, Min_per, ORB_per, AST_per, TO_per,
         twoP_per, drtg, TP_per, dbpm) %>%
  filter(!is.na(yr), !is.na(pts)) %>%
  drop_na() %>%
  filter(
    between(ORB_per, 0, 100),
    between(AST_per, 0, 100),
    between(TO_per, 0, 100),
    between(twoP_per, 0, 100),
    between(TP_per, 0, 100)
  ) %>%
  filter(
    pts >= 0,
    Min_per >= 0,
    drtg >= 0, drtg < 200,
    dbpm < 100
  )

data_clean$yr <- factor(data_clean$yr, levels = c("So", "Jr", "Sr"))
```

# 3. FIT ORIGINAL MODEL

```{r}
model <- lm(
  pts ~ yr + Min_per + ORB_per + AST_per + TO_per +
    twoP_per + drtg + TP_per + dbpm + yr:Min_per,
  data = data_clean
)

summary(model)
```

# 4. ORIGINAL MODEL DIAGNOSTICS

```{r}
e_hat <- residuals(model)
y_hat <- fitted(model)

par(mfrow = c(1,3))

plot(y_hat, data_clean$pts,
     xlab = "Fitted Values",
     ylab = "Points per Game",
     main = "Fitted vs Response")

plot(y_hat, e_hat,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h=0)

qqnorm(e_hat, main="Normal Q-Q Plot")
qqline(e_hat)

par(mfrow=c(1,1))
```

# 5. HISTOGRAMS & DESCRIPTIVE GRAPHS

```{r}
ggplot(data_clean, aes(x = pts)) +
  geom_histogram(
    aes(fill = after_stat(ifelse(xmin == min(xmin), "first", "other"))),
    color = "black",
    bins = 80
  ) +
  scale_fill_manual(values = c("first" = "red", "other" = "grey70")) +
  labs(
    title = "Original Distribution of Points per Game (pts)",
    x = "Points per Game (pts)",
    y = "Count"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

n_obs <- nrow(data_clean)

p1 <- ggplot(data_clean, aes(x = yr)) +
  geom_bar(fill = "grey70", color = "black") +
  labs(title = "Year (yr)") +
  theme_bw()

p2 <- ggplot(data_clean, aes(x = Min_per)) +
  geom_histogram(fill="grey70", color="black", bins=30) +
  labs(title="Minutes Played")

p3 <- ggplot(data_clean, aes(x = ORB_per)) +
  geom_histogram(fill="grey70", color="black", bins=30)

p4 <- ggplot(data_clean, aes(x = AST_per)) +
  geom_histogram(fill="grey70", color="black", bins=30)

p5 <- ggplot(data_clean, aes(x = TO_per)) +
  geom_histogram(fill="grey70", color="black", bins=30)

p6 <- ggplot(data_clean, aes(x = twoP_per)) +
  geom_histogram(fill="grey70", color="black", bins=30)

p7 <- ggplot(data_clean, aes(x = TP_per)) +
  geom_histogram(fill="grey70", color="black", bins=30)

p8 <- ggplot(data_clean, aes(x = drtg)) +
  geom_histogram(fill="grey70", color="black", bins=30)

p9 <- ggplot(data_clean, aes(x = dbpm)) +
  geom_histogram(fill="grey70", color="black", bins=30)

(p1 | p2 | p3) /
  (p4 | p5 | p6) /
  (p7 | p8 | p9)


# Histogram of Min_per (Minutes Percentage) with vertical line at 5
ggplot(data_clean, aes(x = Min_per)) +
  geom_histogram(fill = "grey70", color = "black", bins = 30) +
  geom_vline(xintercept = 5, color = "red", linewidth = 1) +
  labs(
    title = "Histogram of Original Minutes Percentage (Min_per)",
    x = "Minutes Percentage (Min_per)",
    y = "Count"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )
```

# 6. INTERACTION PLOT

```{r}
ggplot(data_clean, aes(x = Min_per, y = pts, color = yr)) +
  geom_smooth(method="lm", se=FALSE, linewidth=1.2)
```

# 7. CORRELATION MATRIX FOR ORIGINAL MODEL

```{r}
numeric_predictors <- data_clean %>%
  dplyr::select(Min_per, ORB_per, AST_per, TO_per,
         twoP_per, drtg, TP_per, dbpm)

cor_matrix <- cor(numeric_predictors, use="pairwise.complete.obs")

corrplot(
  cor_matrix,
  method="color",
  type="upper",
  order="hclust",
  tl.col="black",
  tl.cex=0.8,
  addCoef.col=NA,
  title="Correlation Matrix of Original Model Predictors",
  mar=c(0,0,2,0)
)
```

# 8. CLEAN DATA FOR BOX-COX MODEL

```{r}
data_clean2 <- data %>%
  dplyr::select(
    pts, yr, Min_per,
    treb,
    AST_per, TO_per,
    TS_per,
    dbpm
  ) %>%
  filter(!is.na(yr), !is.na(pts)) %>%
  drop_na() %>%
  filter(
    between(treb, 0, 100),
    between(AST_per, 0, 100),
    between(TO_per, 0, 100),
    between(TS_per, 0, 100)
  ) %>%
  filter(
    pts > 0,
    Min_per > 5,
    dbpm < 100
  ) %>%
  mutate(
    yr = factor(yr, levels = c("So", "Jr", "Sr"))
  )
```

# 9. FIT NON-TRANSFORMED MODEL 2

```{r}
model_final <- lm(
  pts ~ yr + Min_per + treb + AST_per + TO_per +
    TS_per + dbpm + yr:Min_per,
  data = data_clean2
)

summary(model_final)
```

# 10. BOX-COX TRANSFORMATION

```{r}
model_initial <- lm(
  pts ~ yr + Min_per + treb + AST_per + TO_per +
    TS_per + dbpm + yr:Min_per,
  data = data_clean2
)

bc <- boxcox(model_initial, lambda = seq(-2, 2, 0.1))

lambda <- bc$x[which.max(bc$y)]

data_clean2$bc_pts <- (data_clean2$pts^lambda - 1) / lambda
```

# 11. FIT BOX-COX MODEL

```{r}
model_bc <- lm(
  bc_pts ~ yr + Min_per + treb + AST_per + TO_per +
    TS_per + dbpm + yr:Min_per,
  data = data_clean2
)

summary(model_bc)
```

# 12. BOX-COX DIAGNOSTICS

```{r}
# Residuals and fitted values from final BC model
e_hat_final <- residuals(model_bc)
y_hat_final <- fitted(model_bc)

# Layout for 3 diagnostic plots
par(mfrow = c(1, 3))

# 1. Fitted vs Actual
plot(
  y_hat_final, data_clean2$bc_pts,
  xlab = "Fitted Values (BC-transformed scale)",
  ylab = "Observed Points per Game",
  main = "Final Model: Fitted vs Actual"
)

# 2. Residuals vs Fitted
plot(
  y_hat_final, e_hat_final,
  xlab = "Fitted Values (BC-transformed scale)",
  ylab = "Residuals",
  main = "Final Model: Residuals vs Fitted"
)
abline(h = 0)

# 3. QQ Plot
qqnorm(e_hat_final, main = "Final Model: Normal Q-Q Plot")
qqline(e_hat_final)

# Reset layout
par(mfrow = c(1, 1))
```

# 13. PARTIAL F TESTS

```{r}
model_full <- model_bc

model_no_AST <- update(model_full, . ~ . - AST_per)
anova(model_no_AST, model_full)

model_no_TO <- update(model_full, . ~ . - TO_per)
anova(model_no_TO, model_full)

model_no_TS <- update(model_full, . ~ . - TS_per)
anova(model_no_TS, model_full)

model_no_interaction <- update(model_full, . ~ . - yr:Min_per)
anova(model_no_interaction, model_full)

model_no_skills <- lm(
  bc_pts ~ yr + Min_per + dbpm + yr:Min_per,
  data = data_clean2
)
anova(model_no_skills, model_full)

model_min_only <- lm(
  bc_pts ~ Min_per + yr:Min_per,
  data = data_clean2
)
anova(model_min_only, model_full)
```

# 14. INFLUENCE DIAGNOSTICS

```{r}
lev <- hatvalues(model_bc)
rstd <- rstandard(model_bc)
cd <- cooks.distance(model_bc)
dff <- dffits(model_bc)
dfb <- dfbetas(model_bc)

p <- length(coef(model_bc)) - 1
n <- nrow(data_clean2)

cut_lev <- 2*(p+1)/n
cut_r <- 4
cut_cd <- qf(0.5, p+1, n-p-1)
cut_dff <- 2*sqrt((p+1)/n)
cut_dfb <- 2/sqrt(n)

which(lev > cut_lev)
which(abs(rstd) > cut_r)
which(cd > cut_cd)
which(abs(dff) > cut_dff)
apply(abs(dfb) > cut_dfb, 2, which)
```

# 15. MODEL COMPARISON TABLE

```{r}
model1 <- model
model2 <- model_bc

adj_r2_model1 <- summary(model1)$adj.r.squared
adj_r2_model2 <- summary(model2)$adj.r.squared

aic1 <- AIC(model1)
aic2 <- AIC(model2)

bic1 <- BIC(model1)
bic2 <- BIC(model2)

comparison <- data.frame(
  Metric = c("Adjusted R2", "AIC", "BIC"),
  Model1 = c(adj_r2_model1, aic1, bic1),
  Model2 = c(adj_r2_model2, aic2, bic2)
)

comparison
```

# 16. CORRELATION HEATMAP FOR BOX-COX MODEL

```{r}
# Select only variables used in Model 2
data_bc <- data_clean2 %>%
  mutate(bc_pts = (pts^lambda - 1) / lambda)

# Fix namespace conflicts (use dplyr's select explicitly)
data_corr <- data_bc %>%
  dplyr::select(
    bc_pts, yr, Min_per, treb, AST_per, TO_per, TS_per, dbpm
  ) %>%
  mutate(
    yr = as.numeric(factor(yr))
  )

# Compute correlation matrix
corr_mat <- cor(data_corr, use = "pairwise.complete.obs")

# Plot with title
corrplot::corrplot(
  corr_mat,
  method = "color",
  col = colorRampPalette(c("red", "white", "blue"))(200),
  addCoef.col = "black",
  tl.col = "black",
  number.cex = 0.7,
  tl.cex = 0.8,
  title = "Correlation Heatmap for Box-Cox Transformed Model",
  mar = c(0, 0, 2, 0)   # ensures title actually appears
)


```

**SUMMARY**

```{r}
summary(model_bc)
```

# 16. Multicollinearity check

```{r}
vif(model_bc, type = "predictor")
```

# 17. Paired plots for multicollinearity check

```{r}
# Create dataset of only the Box-Cox model predictors
pairs_data <- data_clean2 %>%
  mutate(
    bc_pts = (pts^lambda - 1) / lambda,
    yr_num = as.numeric(yr)   # convert factor to numeric for pairs plot
  ) %>%
  dplyr::select(
    bc_pts,
    yr_num,
    Min_per,
    treb,
    AST_per,
    TO_per,
    TS_per,
    dbpm
  )

# Paired scatterplot matrix
pairs(
  pairs_data,
  main = "Paired Scatterplots for Final Box-Cox Model Predictors",
  pch = 20,
  cex = 0.6
)

```

# 18. Graph for interaction

```{r}
# Create a sequence of Min_per values covering the observed range
newdata <- expand.grid(
  Min_per = seq(min(data_clean2$Min_per), max(data_clean2$Min_per), length.out = 200),
  yr = levels(data_clean2$yr),
  treb = mean(data_clean2$treb),
  AST_per = mean(data_clean2$AST_per),
  TO_per = mean(data_clean2$TO_per),
  TS_per = mean(data_clean2$TS_per),
  dbpm = mean(data_clean2$dbpm)
)

# Add predictions from the Box-Cox model
newdata$pred <- predict(model_bc, newdata)

ggplot(newdata, aes(x = Min_per, y = pred, color = yr)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = "Interaction Effect of Minutes Percentage by Year\n(Box-Cox Model)",
    x = "Minutes Percentage (Min_per)",
    y = "Predicted Box-Cox Transformed Points",
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.border = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```

# 19. Forward Backward Transformation

```{r}
full_formula <- bc_pts ~ yr + Min_per + treb + AST_per + TO_per +
  TS_per + dbpm + yr:Min_per

model_backward <- stepAIC(
  lm(full_formula, data = data_clean2),
  direction = "backward",
  k = 2   # AIC
)
summary(model_backward)

model_forward <- stepAIC(
  lm(bc_pts ~ 1, data = data_clean2),
  scope = list(upper = lm(full_formula, data = data_clean2)),
  direction = "forward",
  k = 2
)
summary(model_forward)

model_stepwise <- stepAIC(
  lm(bc_pts ~ 1, data = data_clean2),
  scope = list(upper = lm(full_formula, data = data_clean2)),
  direction = "both",
  k = 2
)
summary(model_stepwise)

AIC(model_backward, model_forward, model_stepwise)
BIC(model_backward, model_forward, model_stepwise)

```

# 20. Test if Min_per is relevent 

```{r}
model_no_interaction <- update(model_bc, . ~ . - yr:Min_per)

anova(model_no_interaction, model_bc)

model_no_min <- update(model_no_interaction, . ~ . - Min_per)

anova(model_no_interaction, model_no_min)
```


# Split Model 

```{r}
set.seed(67)

data_val <- data_clean2     

n <- nrow(data_val)

train_index <- sample(1:n, size = 0.8*n, replace = FALSE) #80/20 split used

train <- data_val[train_index, ]
test  <- data_val[-train_index, ]
```

# Build model in each dataset

```{r}

# refit final model onto training set
model_train <- lm(
  bc_pts ~ yr + Min_per + treb + AST_per + TO_per + 
    TS_per + dbpm + yr:Min_per, data = train)

summary(model_train)

# refit final model onto testing set
model_test <- lm(bc_pts ~ yr + Min_per + treb + AST_per + TO_per +
    TS_per + dbpm + yr:Min_per,data = test)

summary(model_test)

```


# Compare the two models using VIF

```{r}

#check VIFs

vif(model_train)
vif(model_test)

```

# Check Assumptions

```{r}

# check assumptions in training
par(mfrow=c(3,3))
plot(train$bc_pts ~ fitted(model_train), xlab="fitted", ylab="response", 
     main="Model in Training Data", cex.main=1.5)
plot(residuals(model_train) ~ fitted(model_train), xlab="fitted", ylab="residuals")
for(i in c(2,3,5,6)){
  plot(residuals(model_train) ~ train[[i]], xlab=names(train)[i], ylab="residuals")}
qqnorm(residuals(model_train))
qqline(residuals(model_train))
par(mfrow=c(1,1))

# check assumptions in test
par(mfrow = c(3,3))

plot(test$bc_pts ~ fitted(model_test), xlab="fitted", ylab="response")
title("Model in Test Data", cex.main = 1.5)

plot(residuals(model_test) ~ fitted(model_test), xlab="fitted", ylab="residuals")
for(i in c(2,3,5,6)){
  plot(residuals(model_test) ~ test[[i]], xlab=names(test)[i], ylab="residuals")}

qqnorm(residuals(model_test))
qqline(residuals(model_test))
par(mfrow = c(1,1))

```

# Check Problematic Points

```{r}

# checking problematic observations for train
print("Training Dataset")
# useful values:
n <- nrow(train)
p <- length(coef(model_train))-1
# leverage cutoff
h_cut <- 2*(p+1)/n
print("Leverage Points")
which(hatvalues(model_train) > h_cut)
# outliers
print("Outliers")
which(rstandard(model_train) > 2 | rstandard(model_train) < -2)
# cooks cutoff
print("Influence (Cooks)")
D_cut <- qf(0.5, p+1, n-p-1)
which(cooks.distance(model_train) > D_cut)
# DFFITS cutoff
fits_cut <- 2*sqrt((p+1)/n)
print("Influence (DFFITS)")
which(abs(dffits(model_train)) > fits_cut)
# DFBETAS cutoff
beta_cut <- 2/sqrt(n)
print("Influence (DFBETA)")
for(i in 1:ncol(dfbetas(model_train))){
  print(paste0("Beta ", i-1))
  print(which(abs(dfbetas(model_train)[,i]) > beta_cut))}

# checking problematic observations for test
print("Test Dataset")
# useful values:
n <- nrow(test)
p <- length(coef(model_test))-1
# leverage cutoff
h_cut <- 2*(p+1)/n
print("Leverage Points")
which(hatvalues(model_test) > h_cut)
# outliers
print("Outliers")
which(rstandard(model_test) > 2 | rstandard(model_test) < -2)
# cooks cutoff
print("Influence (Cooks)")
D_cut <- qf(0.5, p+1, n-p-1)
which(cooks.distance(model_test) > D_cut)
# DFFITS cutoff
fits_cut <- 2*sqrt((p+1)/n)
print("Influence (DFFITS)")
which(abs(dffits(model_test)) > fits_cut)
# DFBETAS cutoff
beta_cut <- 2/sqrt(n)
print("Influence (DFBETA)")
for(i in 1:ncol(dfbetas(model_test))){
  print(paste0("Beta ", i-1))
  print(which(abs(dfbetas(model_test)[,i]) > beta_cut))}
  
```

# Summary of Problematic Point Values (this can be taken out if we want, just for me to count)

```{r}

# Summary table of problematic points
cat("TEST SET PROBLEM SUMMARY:\n")
cat("Leverage points:", length(which(hatvalues(model_test) > h_cut)), "/", nrow(test), "\n")
cat("Outliers:", length(which(abs(rstandard(model_test)) > 2)), "/", nrow(test), "\n") 
cat("Cook's D influential:", length(which(cooks.distance(model_test) > D_cut)), "/", nrow(test), "\n")
cat("→ Model stable: no single point dominates\n")

```


# MSE Calculations

```{r}
# Training MSE (in-sample)
train_mse <- mean((train$bc_pts - fitted(model_train))^2)
train_mse

# Test MSE (out-of-sample predictions)
fitted_test <- predict(model_train, newdata = test)
test_mse <- mean((test$bc_pts - fitted_test)^2)
test_mse

#Test-Train Ratio
test_mse/train_mse

```

